Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-08-16 src/tex/ms.tex
mention new Selsis et al. results

Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, \rev{the atmospheric composition~\citep{2020NatGe..13..265G} and comprehensive analysis of meteoritic samples across the Solar System~\citep{2018SSRv..214...36A,2022Natur.611..245B}} suggest that it went through this phase.
Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.

@@

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
With even a fraction of the Earth's water inventory, a planet absorbing more flux than the radiation limit of steam atmospheres will enter a runaway greenhouse state resulting in a global magma ocean~\citep{Lichtenberg2021c,Boukrouche2021}.
We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the distribution of water between planetary interior and atmosphere, and on the resulting thermal atmospheric structure~\citep{Dorn2021,Salvador2023}.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
However, already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard expected outcome~\citep[e.g.,][]{2019PNAS..116.9723Z,Venturini2020,Emsenhuber2021b,Schlecker2021,Burn2021}.
But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system\rev{~\citep{McCubbin2019,2020Sci...369.1110P,2021PSJ.....2..244J}} suggests that hydrogen is present in all rocky planets upon formation.
Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
Enhanced equilibration between the core, mantle, and atmosphere during magma ocean evolution of rocky exoplanets further enhances this process~\citep{2021ApJ...914L...4L,Schlichting2022}.
Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution.
Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M~dwarf systems~\citep{2022ApJ...938L...3L}.

>>>>

Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage. % if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, \rev{the atmospheric composition~\citep{2020NatGe..13..265G} and comprehensive analysis of meteoritic samples across the Solar System~\citep{2018SSRv..214...36A,2022Natur.611..245B}} suggest that it went through this phase.
Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.

@@

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
With even a fraction of the Earth's water inventory, a planet absorbing more flux than the radiation limit of steam atmospheres \revii{was found to} enter a runaway greenhouse state resulting in a global magma ocean~\revii{\citep[][but see \citet{Selsis2023} for a contrasting viewpoint]{Lichtenberg2021c,Boukrouche2021}}.
We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the distribution of water between planetary interior and atmosphere, and on the resulting thermal atmospheric structure~\citep{Dorn2021,Salvador2023}.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
\revii{It was shown that} already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
\revii{However, recent simulations accounting for radiative–convective profiles suggest the existence of cooler runaway greenhouse states that do not necessarily yield a surface magma ocean~\citep{Selsis2023}.
Climate state and atmospheric structure, as well as the fraction of planets fulfilling the requirements of a steam atmosphere} have an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard expected outcome~\citep[e.g.,][]{2019PNAS..116.9723Z,Venturini2020,Emsenhuber2021b,Schlecker2021,Burn2021}.
But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system\rev{~\citep{McCubbin2019,2020Sci...369.1110P,2021PSJ.....2..244J}} suggests that hydrogen is present in all rocky planets upon formation.
Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
Enhanced equilibration between the core, mantle, and atmosphere during magma ocean evolution of rocky exoplanets further enhances this process~\citep{2021ApJ...914L...4L,Schlichting2022}.
Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution.
Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M~dwarf systems~\citep{2022ApJ...938L...3L}.
